Thesis Title

### 2022-09-01 chapters/discussion.tex
improvements to discussion

The constant changes in versions and/or the discontinuation of a bioinformatics tool complicates the standardisation of data analysis. In routine settings, automation and standardisation of the analysis are significant for the reliability of the diagnostic test results. With the lack of proper benchmarks to validate in what approach is to be followed, no hope of standardisation can be achieved.

@@

In Chapter \ref{ch:paper1}, the suitability of SMg for the microbiological diagnosis was assessed, with particular emphasis on the bioinformatic process. In total, 3 different bioinformatics pipelines were evaluated to identify those which could provide a clinical microbiologist with the maximum of relevant and accurate information: an open-source Unix-based approach, and two commercial alternatives, Basespace and CLC Genomics Workbench. No approach outperformed the other, and very disparate results were obtained for each one. One of the limitations of this study was that, although several approaches were used, no comprehensive assessment of the tools available for each step of the workflow was performed (quality control, removal of host sequences, taxonomic identification, gene detection and \textit{de novo} assembly). Overall, we found that CLC Genomics Workbench has advantages over the other methods as it does not require previous knowledge of Unix-based tools, it is arguably the most user-friendly, but this conclusion is limited to the set of tools evaluated. Currently, no open-source, user-friendly, one-stop, comprehensive metagenomics toolkit with a visual user interface for shotgun metagenomics analysis is available, but efforts are being made to change this paradigm. Alternatives, such as Anvi'o \citep{eren_anvio_2015} for shotgun metagenomics, and Qiime2 \citep{bolyen_reproducible_2019} for metataxonomics, leverage the individual components into a streamlined manner, but require command-line familiarity. Still, each individual component of such software should be individually evaluated.

In Chapter \ref{ch:paper5}, 12 \textit{de novo} assemblers were benchmarked with the same metagenomic samples, selected based on the date of last update. No assembler stood out as an undisputed all-purpose choice for short-read metagenomic prokaryote \textit{de novo} genome assembly, highlighting that efforts are still needed to further improve metagenomic assembler performance. This arose directly from the need to pick an assembler to implement in Chapter \ref{ch:paper4}, where ultimately a two-pronged approach was followed including both a metagenomic (MEGAHIT) and a traditional (SPAdes) genomic assembler.

Despite these efforts, just a small subset of the commonly bioinformatics methods applied to \ac{SMg} have been assessed. Regarding the binning process, disparate results have been observed \citep{bharti_current_2021, yue_evaluating_2020, yang_review_2021, sczyrba_critical_2017}, highlighting that the need for proper benchmark cannot be limited to the tools itself, but also the standardisation of the data used for the benchmark. Similarly, the same has been observed for taxonomic assignment \citep{sczyrba_critical_2017,ye_benchmarking_2019, tamames_assessing_2019}.

@@

In Chapter \ref{ch:paper5}, its shown that suitable mock communities, reproducing the users’ samples of interest, can be used as a gold standard to evaluate tool performance. Several well characterised mock communities are currently available to be used by the community when assessing and benchmarking software suitable for \ac{SMg} analysis. Chapter \ref{ch:paper5} features the ZymoBIOMICS Microbial Community Standard, composed of eight bacterial species and two fungi, and is available commercially, with reference sequences made publicly available, with both even and logarithmically distribution of species\footnote{\url{https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards}}. Still, ZymoBIOMICS Microbial Community Standards might not be representative of the metagenomic complexity of the samples of interest of most researchers, its relative simplicity means that the results shown probably represent a best-case scenario.

Alternatively, more complex communities, with good metadata and reference genomes publicly available, are available for acquisition and sequencing. \ac{ATCC} 10 Strain Even Mix Genomic Material
MSA-1000\footnote{\url{https://www.atcc.org/products/msa-1000}} and 20 Strain Even Mix Genomic Material MSA-1002\footnote{\url{https://www.atcc.org/products/msa-1002}} consists of, respectively, 10 and 20, fully sequenced, characterised, and authenticated \ac{ATCC} Genuine Cultures mixed evenly, selected based on pathogen relevance. It also provides Gut Microbiome Genomic Mix MSA-1006\footnote{\url{https://www.atcc.org/products/msa-1006}} composed of an even mixture of genomic DNA prepared from 12 fully sequenced, characterised, and authenticated bacterial species observed in normal and atypical gut microbial communities.

For the replication of more complex communities, the MICROBIOME Community of Special Interest\footnote{\url{https://www.microbiome-cosi.org/}} makes available, through their \ac{CAMI} initiative\footnote{\url{https://data.cami-challenge.org/}}, several datasets varying in complexity, replicating several communities such as human microbiome and the rhizosphere. Sources of truth are not made available, including the complete genomes of the community profile, so comparisons are difficult.

@@

One of the biggest challenges when dealing with metagenomic data is the lack of golden standards, although major efforts are being made on the standardisation and assessment of software, both commercial and open source \citep{angers-loustau_challenges_2018, gruening_recommendations_2019, sczyrba_critical_2017, couto_critical_2018}. A plethora of free-to-use, open-source tools are available specifically for metagenomic data, both short and long read data, and several combinations of these tools can be used to characterise the causative agent in a patient's infection in a fraction of time required by the traditional methods. Unfortunately, as observed in Chapter \ref{ch:paper5}, massive differences in success and performance were observed in software built for the same purpose, in this case the \textit{de novo} assembly of metagenomic genomic data. There are several steps that can be implemented to ensure the transparency and reproducibility of the chosen workflow for metagenomic analysis, regardless of the tools chosen.

>>>>

The constant changes in versions and/or the discontinuation of a bioinformatics tool complicates the standardisation of data analysis. In routine settings, automation and standardisation of the analysis are significant for the reliability of the diagnostic test results. With the lack of proper benchmarks to validate what approach is to be followed, no hope of standardisation can be achieved.

@@

In Chapter \ref{ch:paper1}, the suitability of \ac{SMg} for the microbiological diagnosis was assessed, with particular emphasis on the bioinformatic process. In total, 3 different bioinformatics pipelines were evaluated to identify those which could provide a clinical microbiologist with the maximum of relevant and accurate information: an open-source Unix-based approach, and two commercial alternatives, Basespace and CLC Genomics Workbench. No approach outperformed the other, and very disparate results were obtained for each one. One of the limitations of this study was that, although several approaches were used, no comprehensive assessment of the tools available for each step of the workflow was performed (quality control, removal of host sequences, taxonomic identification, gene detection and \textit{de novo} assembly). Currently, no open-source, user-friendly, one-stop, comprehensive metagenomics toolkit with a visual user interface for shotgun metagenomics analysis is available, but efforts are being made to change this paradigm. Alternatives, such as Anvi'o \citep{eren_anvio_2015} for shotgun metagenomics, and Qiime2 \citep{bolyen_reproducible_2019} for metataxonomics, leverage the individual components in a streamlined manner, but require command-line familiarity. Still, each individual component of such software should be individually evaluated.

In Chapter \ref{ch:paper4}, where ultimately a two-pronged approach was followed including both a genomic and a metagenomic \textit{de novo} assembler. This arose from the need to recover complete CDS sequences, where none of the alternatives on their own was able to perform consistently. Despite several benchmarking studies being available \citep{miller_assembly_2010, earl_assemblathon_2011, bradnam_assemblathon_2013, liao_current_2019, chen_benchmarking_2020, sczyrba_critical_2017, meyer_critical_2022}, the assemblers benchmarked and the datasets used are not consistent, usually limited to a couple of different tools that do not reflect the current landscape of possible \textit{de novo} assembly software. These benchmarks are also usually coupled with the release of a new tool, therefore often skewed towards certain results and do not offer an unbiased assessment.

In Chapter \ref{ch:paper5}, 12 \textit{de novo} assemblers were benchmarked with the same metagenomic samples, selected based on the date of the last update to the software and with a cut-off in 2015. No assembler stood out as an undisputed all-purpose choice for short-read metagenomic prokaryote \textit{de novo} genome assembly, highlighting that efforts are still needed to further improve metagenomic assembler performance.

Despite these efforts, just a small subset of the common bioinformatics methods applied to \ac{SMg} have been assessed. Regarding the binning process, disparate results have been observed \citep{bharti_current_2021, yue_evaluating_2020, yang_review_2021, sczyrba_critical_2017}, highlighting that the need for proper benchmark cannot be limited to the tools itself, but also the standardisation of the data used for the benchmark. Similarly, the same has been observed for taxonomic assignment \citep{sczyrba_critical_2017,ye_benchmarking_2019, tamames_assessing_2019}.

@@

Chapter \ref{ch:paper5} shows that suitable mock communities, reproducing the users’ samples of interest, can be used as a gold standard to evaluate tool performance. Several well-characterised mock communities are currently available to be used by the community when assessing and benchmarking software suitable for \ac{SMg} analysis. Chapter \ref{ch:paper5} features the ZymoBIOMICS Microbial Community Standard, composed of eight bacterial species and two fungi, and is available commercially, with reference sequences made publicly available, with both even and logarithmically distribution of species\footnote{\url{https://www.zymoresearch.com/collections/zymobiomics-microbial-community-standards}}. Still, ZymoBIOMICS Microbial Community Standards might not be representative of the metagenomic complexity of the samples of interest of most researchers, its relative simplicity means that the results shown probably represent a best-case scenario. Furthermore, in the logarithmically distributed sample, six replicons have a coverage lower than one time, rendering it unrecoverable through assembly. To address this, two other community standards were analysed: the BMock12 \citep{sevim_shotgun_2019} and the Gut-Mix-RR \citep{amos_developing_2020}.

The BMock12 community is composed of twelve bacterial species, spanning eight genera. This community dataset includes several closely related strains: two replicons of \textit{Halomonas} sp., three replicons of the \textit{Micromonospora} and two replicons of \textit{Marinobacter} sp. The challenges of this data set are to assemble the genomes of the two Marinobacter and the two Halomonas strains, which come at 85\% and 99\% ANIb, respectively. Samples were made available of the sequencing products of this community both in second (Illumina) and third-generation sequencing (PacBio and Oxford Nanopore), expanding its utility for the benchmarking of both short-read, long-read and hybrid software. With worth noting that the molarity of each genome in the mock community is not even throughout, resulting in the extreme overrepresentation of some taxa in relation to others. This can have the benefit of providing a form of a limit of detection assessment, depending on the type of analysis performed. Still, besides the challenges that this dataset present, is still a rather simplistic picture of a metagenomic sample.

The Gut-Mix-RR community dataset, and its uneven distribution of species counterpart Gut-HiLo-RR, spans five phyla, thirteen families, sixteen genera and nineteen species, comprising twenty strains that aim to emulate the healthy human gut microbiota. The strains were obtained from Leibniz Institute DSMZ-German Collection of Microorganisms and Cell Cultures GmbH. The genomes in the mock community are fairly diverse, with the challenging genomes being the two \textit{Bifidobacterium longum} subspecies (ANIb=0.95). It’s worth denoting that no complete genome is available for eight of the genomes, including one of the \textit{Bifidobacterium longum} subspecies. This makes the assessment of results challenging, if not impossible, as no source of absolute truth is available.

For the replication of more complex communities, the MICROBIOME Community of Special Interest\footnote{\url{https://www.microbiome-cosi.org/}} makes available, through their \ac{CAMI} initiative\footnote{\url{https://data.cami-challenge.org/}}, several datasets varying in complexity, replicating several communities such as human microbiome and the rhizosphere \citep{sczyrba_critical_2017, meyer_critical_2022}. Unfortunately, sources of truth are not made available, including the complete genomes of the community profile, and assessment of results is difficult, if not impossible. Despite their popularity, no information is provided in the manuscripts or in the supplemental materials about the source of the genomes in each dataset. It is only mentioned that the community composition was designed according to specified criteria. Additionally, only assemblies are made available, and not the source references for the generation of the mock datasets. These assemblies are, unfortunately, often the result of a pooled assembly and are extremely fragmented, with N50 ranging from 18406 basepairs in 56168 contigs to 58988 basepairs in 508968 contigs. This makes this dataset not suitable for benchmark studies, which require defined mock communities as a source of truth.

The Fishing in the Soup dataset \citep{grutzke_fishing_2019} is a mock community containing DNA of foodborne bacteria. The standard consists of 34 equimolarly pooled bacterial DNAs belonging to 17 genera and 30 species, elected based on their incidence in foodstuff and pathogenicity. Despite its diversity, not all reference sequences are available, and the used reference sequences are not the ones corresponding to the sequenced strains in some cases. Although two sets of the same species replicons are present (\textit{Bacillus cereus}, \textit{Salmonella enterica}, \textit{Clostridium perfringens}), their ANIb is lower than the BMock12 \textit{halomonas} replicons, and no independent references for different strains of the same species are provided.

Similarly to the ZymoBIOMICS Microbial Community Standard, communities with good metadata and reference genomes publicly available are available for acquisition and sequencing from \ac{ATCC}. The \ac{ATCC} 10 Strain Even Mix Genomic
Material MSA-1000\footnote{\url{https://www.atcc.org/products/msa-1000}} and 20 Strain Even Mix Genomic Material MSA-1002\footnote{\url{https://www.atcc.org/products/msa-1002}} consists of, respectively, 10 and 20, fully sequenced, characterised, and authenticated \ac{ATCC} Genuine Cultures mixed evenly, selected based on pathogen relevance. It also provides Gut Microbiome Genomic Mix MSA-1006\footnote{\url{https://www.atcc.org/products/msa-1006}} composed of an even mixture of genomic DNA prepared from 12 fully sequenced, characterised, and authenticated bacterial species observed in normal and atypical gut microbial communities. Just like ZymoBIOMICS Microbial Community Standards, the simplicity of the ATCC mock communities might not be representative of the metagenomic complexity of the samples of interest to most researchers, representing simplistic scenarios with limited usability.

No ideal solution is currently available in the research community or the market, highlighting the need for better, more diverse and well-described community standards representing realistic scenarios.

@@

One of the biggest challenges when dealing with metagenomic data is the lack of golden standards, although major efforts are being made on the standardisation and assessment of software, both commercial and open source \citep{angers-loustau_challenges_2018, gruening_recommendations_2019, sczyrba_critical_2017, couto_critical_2018}. A plethora of free-to-use, open-source tools are available specifically for metagenomic data, both short and long-read data, and several combinations of these tools can be used to characterise the causative agent in a patient's infection in a fraction of time required by the traditional methods. Unfortunately, as observed in Chapter \ref{ch:paper5}, massive differences in success and performance were observed in software built for the same purpose, in this case the \textit{de novo} assembly of metagenomic genomic data. There are several steps that can be implemented to ensure the transparency and reproducibility of the chosen workflow for metagenomic analysis, regardless of the tools chosen.
